Categorical Foundations of E-Graphs With Binding

### 2025-07-10 LICS_2025/e_graphs_with_bindings/sections/appendix.tex
one more bib fix

    This construction is analogous to how a pushout was constructed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5.5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-2.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-1.5, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

\begin{figure}
\[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/dpo_example}
}
\]
\caption{$\beta$-rule DPOI example.}
\label{fig:dpoi-example}
\end{figure}

>>>>

    This construction is analogous to how a pushout was constructed in~\cite{tiurin2025equivalencehypergraphsdporewriting} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5.5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-2.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-1.5, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

\begin{figure}
    \ifdefined\ONECOLUMN
    \[
    \adjustbox{width=0.6\linewidth}{
    \tikzfig{./figures/dpo_example}
    }
    \]
    \else
    \[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/dpo_example}
}
\]
    \fi
    \caption{$\beta$-rule DPOI example.}
    \label{fig:dpoi-example}
\end{figure}

### 2025-07-10 LICS_2025/e_graphs_with_bindings/sections/categorical_semantics.tex
one more bib fix

To encode equivalences between morphisms (or $\Sigma$-terms), the hom-sets of the corresponding free categories can be endowed with the additional structure of a semilattice induced by an idempotent, commutative and associative operator \textit{join} $+$ as presented in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
This endowment turns such categories into $\catname{SLat}$-enriched categories where $\catname{SLat}$ is a category of \textit{unbounded} semilattices and homomorphisms between them (see Appendix~\ref{sec:appendix:slat}).

@@

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Figure \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.
Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.
% Figure \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

>>>>

To encode equivalences between morphisms (or $\Sigma$-terms), the hom-sets of the corresponding free categories can be endowed with the additional structure of a semilattice induced by an idempotent, commutative and associative operator \textit{join} $+$ as presented in~\cite{tiurin2025equivalencehypergraphsdporewriting}.
This endowment turns such categories into $\catname{SLat}$-enriched categories where $\catname{SLat}$ is a category of \textit{unbounded} semilattices and homomorphisms between them (see Appendix~\ref{sec:appendix:slat}).

@@

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Fig. \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{tiurin2025equivalencehypergraphsdporewriting}; the first component denotes an empty diagram.
Note the orientation of string diagrams are now from top to bottom as opposed to what was used in~\nameref{sec:introduction}: the latter was mimicking the orientation of e-graphs, \textit{i.e.}, traditional orientation for expression trees.
% Fig. \ref{fig:string-equations} displays the additional equations which these diagrams satisfy, in addition to the standard SMC equations. 
% The first four equations are those displayed in Definition \ref{def:enriched-prop},  while the final four axiomatize $+$ as an \textit{n-ary} associative, commutative and idempotent operation.  We overload the binary notation $+$ for our $n$-ary notation.  
% We will later prove the intuitive fact that these diagrams are sound and complete with respect to their intended categorical semantics, noting that similar diagrammatic languages using boxes to express choice have been used before~\cite{duncan_generalised_2009}.

### 2025-07-10 LICS_2025/e_graphs_with_bindings/sections/combinatorial_semantics.tex
one more bib fix

\section{Closed E-hypergraphs}
Prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established body of literature on hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

We follow an analogous approach by combining combinatorial representations of~\cite{fscd} and~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
We next extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} to accommodate new structure brought by closedness, and when evoking a term `e-hypergraph' we will mean the structure defined as follows.

@@

Intuitively, edges from $\textcolor{e-color}{E}$ encode equivalence classes, \emph{i.e.}, e-boxes, while edges from $\textcolor{closed-color}{E}$ encode lambda-abstraction boxes.
Also note the typing on the $l_{V}$ function.
We work with freely generated closed semilattice-enriched categories, and to avoid unnecessary complications we limit the labelling of vertices to such a type.
In particular, it prohibits vertices to be labelled with, \emph{e.g.}, $A \otimes B$, but allows them to be labelled with $A \otimes B \multimap C \otimes D$.
$\textcolor{closed-color}{<}$ defines closed monoidal structure as defined in~\cite{fscd} and $(\textcolor{e-color}{<},\consistency)$ defines $\catname{SLat}$ structure as per~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

@@

Because extended cospans have a more general notion of interface,  including \textit{internal} vertices,  DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics} needs some adjustments, defined in this section.
The construction in this section closely follows the one introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} with some adjustments to account for $\textcolor{closed-color}{<}$.
We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogously to Section \ref{sec:combinatorial-semantics}, observing that  cospan
$
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m$ expresses the same data as $
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
$
allows us to encode rewrite rules as extended cospans of the form
$
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R},
$ 
which fits into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.

>>>>

\section{Closed E-hypergraphs}
Prior work of~\cite{tiurin2025equivalencehypergraphsdporewriting} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established body of literature on hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{tiurin2025equivalencehypergraphsdporewriting} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

We follow an analogous approach by combining combinatorial representations of~\cite{fscd} and~\cite{tiurin2025equivalencehypergraphsdporewriting}.
We next extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{tiurin2025equivalencehypergraphsdporewriting} to accommodate new structure brought by closedness, and when evoking a term `e-hypergraph' we will mean the structure defined as follows.

@@

Intuitively, edges from $\textcolor{e-color}{E}$ encode equivalence classes, \emph{i.e.}, e-boxes, while edges from $\textcolor{closed-color}{E}$ encode lambda-abstraction boxes.
Also note the typing on the $l_{V}$ function.
We work with freely generated closed semilattice-enriched categories, and to avoid unnecessary complications we limit the labelling of vertices to such a type.
In particular, it prohibits vertices to be labelled with, \emph{e.g.}, $A \otimes B$, but allows them to be labelled with $A \otimes B \multimap C \otimes D$.
$\textcolor{closed-color}{<}$ defines closed monoidal structure as defined in~\cite{fscd} and $(\textcolor{e-color}{<},\consistency)$ defines $\catname{SLat}$ structure as per~\cite{tiurin2025equivalencehypergraphsdporewriting}.

@@

Because extended cospans have a more general notion of interface,  including \textit{internal} vertices,  DPOI rewriting as presented in Section \ref{sec:combinatorial-semantics} needs some adjustments, defined in this section.
The construction in this section closely follows the one introduced in~\cite{tiurin2025equivalencehypergraphsdporewriting} with some adjustments to account for $\textcolor{closed-color}{<}$.
We do not expect internal interfaces to be preserved during rewriting: for example,  when the semilattice equations are considered as rewrites. 
Thus,  we wish a rewrite rule to be a pair of \textit{extended} cospans of e-hypergraphs with matching \textit{external} (but not necessarily internal) interfaces,  as follows. 
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{L} \xleftarrow{} m' \xleftarrow{} m
\qquad
    n \xrightarrow{} n'' \xrightarrow{} \mathcal{R} \xleftarrow{} m'' \xleftarrow{} m
\] 
Analogously to Section \ref{sec:combinatorial-semantics}, observing that  cospan
$
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{L} \xleftarrow{} n'' + m'' \xleftarrow{} n + m$ expresses the same data as $
    0 \xrightarrow{} 0 \xrightarrow{} \mathcal{R} \xleftarrow{} n' + m' \xleftarrow{} n + m
$
allows us to encode rewrite rules as extended cospans of the form
$
\mathcal{L} \xleftarrow{} n' + m' \xleftarrow{} n + m \xrightarrow{} n'' + m'' \xrightarrow{} \mathcal{R},
$ 
which fits into the DPO formalism. We make this identification throughout,  without further mention. 
%We will futher omit the arrows $0 \to 0 \to \mathcal{L}$.

### 2025-07-10 LICS_2025/e_graphs_with_bindings/sections/introduction.tex
one more bib fix

All these issues can be resolved by the use of \emph{string diagrams} in place of textual syntax.
\cite{ghica2024equivalencehypergraphsegraphsmonoidal}~shows how e-graphs (without binding) can be given a categorical semantics using semilattice-enriched symmetric monoidal categories, which allows e-graphs to be reconstructed as string diagrams.
In this paper, we extend this technique to e-graphs with bindings, lifting to monoidal \emph{closed} categories.

@@

The general background literature for string diagrams is summarised in~\cite{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams with bindings is given in~\cite{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

@@

\begin{figure*}[h!]
\[
\adjustbox{scale=0.6}{
\tikzfig{figures/e-graph-example}
}
\]
\caption{E-graph example (top) and its equivalent string diagram representation (bottom)~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}}
\label{fig:e-graph-example}
\end{figure*}

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the approach in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}. 
Note that for the initial term, $(a*2)/2$ (subfigure (a)), which is already represented efficiently in the e-graph by sharing the node `2', the string diagram version shares this node explicitly, using a contraction node.

>>>>

All these issues can be resolved by the use of \emph{string diagrams} in place of textual syntax.
\cite{tiurin2025equivalencehypergraphsdporewriting}~shows how e-graphs (without binding) can be given a categorical semantics using semilattice-enriched symmetric monoidal categories, which allows e-graphs to be reconstructed as string diagrams.
In this paper, we extend this technique to e-graphs with bindings, lifting to monoidal \emph{closed} categories.

@@

The general background literature for string diagrams is summarised in~\cite{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams with bindings is given in~\cite{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\cite{tiurin2025equivalencehypergraphsdporewriting}.

@@

\begin{figure*}[h!]
\[
\adjustbox{scale=0.6}{
\tikzfig{figures/e-graph-example}
}
\]
\caption{E-graph example (top) and its equivalent string diagram representation (bottom)~\cite{tiurin2025equivalencehypergraphsdporewriting}}
\label{fig:e-graph-example}
\end{figure*}

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the approach in~\cite{tiurin2025equivalencehypergraphsdporewriting}. 
Note that for the initial term, $(a*2)/2$ (subfigure (a)), which is already represented efficiently in the e-graph by sharing the node `2', the string diagram version shares this node explicitly, using a contraction node.

### 2025-07-10 POPL2026/paper/sections/appendix.tex
one more bib fix

    This construction is analogous to how a pushout was constructed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
    \adjustbox{scale=0.8}{
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-4, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-3, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-2.5, 3.5) {};
	\end{pgfonlayer}
    }
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

>>>>

    This construction is analogous to how a pushout was constructed in~\cite{tiurin2025equivalencehypergraphsdporewriting} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
    \adjustbox{scale=0.8}{
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-4, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-3, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-2.5, 3.5) {};
	\end{pgfonlayer}
    }
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

### 2025-07-10 POPL2026/paper/sections/categorical_semantics.tex
one more bib fix

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Fig. \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}; the first component denotes an empty diagram.

@@

	\noindent
	plus the rules for equivalence of terms up to renaming of slots.
	The earlier work of~\citet{ghica2024equivalencehypergraphsegraphsmonoidal} that introduced free semilattice enrichment to describe e-graphs using morphism terms used the $+$ of a semilattice to denote the equivalence of morphism subterms.
	For example, the e-graph (b) from \autoref{fig:e-graph-example} can be written as a term as 
	\[(a \otimes ((2;\comonoid) \otimes (1;id));\sym);((* \otimes \counit + (id \otimes \counit \otimes id);<\!\!<) \otimes id);/~,\] where $+$ encodes the equivalence of $*$ and $<\!\!<$
	We can see that the first three rules (plus reflexivity, symmetry, and transitivity) above correspond to idempotence, commutativity and associativity of $+$ and the fact that $+$ respect the composition $(;)$.
	%  and tensoring $\otimes$ (traditional e-graphs do not have tensoring as they are defined for algebraic signatures).
	Then, while the binding congruence rule is \textit{natural} to enforce when developing e-graphs with binding, we see that it is indeed the \textit{right} rule to enforce as we can see it follow by a purely categorical argument by the free enrichment as per the law \ref{law:distributivity}.
	We do not enforce any rules for renaming as string diagrams are essentially a nameless representation of terms.
	Moreover, some renaming rules are not consistent with string diagrams, as slotted e-graph try to represent, \textit{e.g.}, terms $(a + b)$ and $(c + d)$ as a single entity (up to renaming of free variables), which in the case for string diagrams require an additional step of closure conversion.

>>>>

To aid reasoning, we introduce a new language of string diagrams for closed $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms as well as another hierarchical ``box'' for lambda abstraction.
It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting and in interpreting $\lambda$-terms.
Fig. \ref{fig:egraph-strings} displays the generators of this language which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for lambda abstraction as established in~\cite{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice-enrichment as introduced in~\cite{tiurin2025equivalencehypergraphsdporewriting}; the first component denotes an empty diagram.

@@

	\noindent
	plus the rules for equivalence of terms up to renaming of slots.
	The earlier work of~\citet{tiurin2025equivalencehypergraphsdporewriting} that introduced free semilattice enrichment to describe e-graphs using morphism terms used the $+$ of a semilattice to denote the equivalence of morphism subterms.
	For example, the e-graph (b) from \autoref{fig:e-graph-example} can be written as a term as 
	\[(a \otimes ((2;\comonoid) \otimes (1;id));\sym);((* \otimes \counit + (id \otimes \counit \otimes id);<\!\!<) \otimes id);/~,\] where $+$ encodes the equivalence of $*$ and $<\!\!<$
	We can see that the first three rules (plus reflexivity, symmetry, and transitivity) above correspond to idempotence, commutativity and associativity of $+$ and the fact that $+$ respect the composition $(;)$.
	%  and tensoring $\otimes$ (traditional e-graphs do not have tensoring as they are defined for algebraic signatures).
	Then, while the binding congruence rule is \textit{natural} to enforce when developing e-graphs with binding, we see that it is indeed the \textit{right} rule to enforce as we can see it follow by a purely categorical argument by the free enrichment as per the law \ref{law:distributivity}.
	We do not enforce any rules for renaming as string diagrams are essentially a nameless representation of terms.
	Moreover, some renaming rules are not consistent with string diagrams, as slotted e-graph try to represent, \textit{e.g.}, terms $(a + b)$ and $(c + d)$ as a single entity (up to renaming of free variables), which in the case for string diagrams require an additional step of closure conversion.

### 2025-07-10 POPL2026/paper/sections/combinatorial_semantics.tex
one more bib fix

\section{Closed E-hypergraphs}
\label{sec:closed-e-hyp}
Prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established body of literature on hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

We follow an analogous approach by combining combinatorial representations of~\cite{fscd} and~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.
We next extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} to accommodate new structure brought by closedness, and when evoking a term `e-hypergraph' we will mean the structure defined as follows.

>>>>

\section{Closed E-hypergraphs}
\label{sec:closed-e-hyp}
Prior work of~\cite{tiurin2025equivalencehypergraphsdporewriting} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established body of literature on hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{tiurin2025equivalencehypergraphsdporewriting} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

We follow an analogous approach by combining combinatorial representations of~\cite{fscd} and~\cite{tiurin2025equivalencehypergraphsdporewriting}.
We next extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{tiurin2025equivalencehypergraphsdporewriting} to accommodate new structure brought by closedness, and when evoking a term `e-hypergraph' we will mean the structure defined as follows.
